---
title: "Hayman Fire Recovery"
author: "Meggi Varnai"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document: default
  word_document: default
  pdf_document: default
  
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_file = 'index',
      output_dir='../'
    )
  })
---


```{r setup, warning=F,message=F, include=FALSE}
library(tidyverse)
library(tidyr)
library(ggthemes)
library(lubridate)
# Now that we have learned how to munge (manipulate) data
# and plot it, we will work on using these skills in new ways
knitr::opts_knit$set(root.dir='..')
```


```{r dataread, warning=F,message=F}
####-----Reading in Data and Stacking it ----- ####
#Reading in files
files <- list.files('data',full.names=T)
#Read in individual data files
ndmi <- read_csv(files[1]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndmi')
ndsi <- read_csv(files[2]) %>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndsi')
ndvi <- read_csv(files[3])%>% 
  rename(burned=2,unburned=3) %>%
  mutate(data='ndvi')
# Stack as a tidy dataset
full_long <- rbind(ndvi,ndmi,ndsi) %>%
  gather(key='site',value='value',-DateTime,-data) %>%
  filter(!is.na(value))
```



## Question 1) 

What is the correlation between NDVI and NDMI? - here I want you to
convert the full_long dataset in to a wide dataset using the 
function "spread" and then make a plot that shows the correlation as a
function of if the site was burned or not (x axis should be ndmi)
You should exclude winter months and focus on summer months

```{r message=F, warning=F,fig.width=7,fig.height=7}

full_wide <- spread(full_long,key='data',value='value') %>%
  filter_if(is.numeric,all_vars(!is.na(.))) %>%
  mutate(month=month(DateTime),
         year=year(DateTime))

#create variable for summer months
var.summer_months<- c(5,6,7,8)

summer<-full_wide %>%
  filter(month %in% var.summer_months)

ggplot(summer,aes(x=ndmi,y=ndvi,color=site))+
  geom_point()+
  xlim(-0.5,0.5)+
  theme_few()+
  theme(legend.position=c(0.2,0.8))

#correlation
cor(summer$ndmi,summer$ndvi)
  
```

A: Based off the plot, there is a positive correlation between ndmi and ndvi. I ran a correlation test and found that the correlation for summer months between NDMI and NDVI is 0.68. This initial analysis is not considering any other variables (ie burned vs unburned), which would be worth looking into to determine if there are stronger relationships present. 

## Question 2 

2) What is the correlation between average NDSI (normalized
 snow index) for January - April and average NDVI for June-August?
In other words, does the previous year's snow cover influence vegetation growth for the following summer?

```{r warning=F,message=F,fig.width=7,fig.height=7}
#variable ndsi months
var.snow_months<-c(1,2,3,4)

#variable ndvi months
var.growth_months<-c(6,7,8)

#mean NDSI per year
ndsi_avg<-full_wide[c("DateTime","ndsi","month","year","site")] %>%
  filter(month %in% var.snow_months) %>% 
  group_by(site,year) %>%
  summarize(ndsi_avg=mean(ndsi))


#mean NDVI per year
ndvi_avg<-full_wide[c("DateTime","ndvi","month","year","site")] %>%
  filter(month %in% var.growth_months) %>%
  group_by(site,year) %>%
  summarize(ndvi_avg=mean(ndvi))


#combining NDVI and NDSI into one dataset
combined<-inner_join(ndvi_avg,ndsi_avg) 

#correlation
cor(combined$ndvi_avg,combined$ndsi_avg)

#plot
ggplot(combined, aes(x=ndvi_avg, y=ndsi_avg,color=site))+
  geom_point()+
  theme_few()+
  theme(legend.position=c(0.2,0.8))

```
When we are considering all sites and time frames, the data shows that there is low correlation between snow cover and vegetation growth, 0.18. This low correlation between the previous year's snow cover on vegetation growth for the following summer excludes several variables like burned and unburned, which on our plot we can see how that there is a difference between these variables. Our analysis also doesn't account for differences in runoff patterns due to elevation and slope that would impact the correlation between snow cover and vegetation growth. 

## Q3

How is the snow effect from question 2 different between pre- and post-burn
and burned and unburned? 

```{r ,message=F,fig.width=7,fig.height=7}
#create dataframe on pre-and post 2002- hayman fire.
preyears<-c(1984:2002)
postyears<-c(2003:2019)

#Snow
presnow<-ndsi_avg %>%
  filter(year %in% preyears) %>%
  group_by(year) %>%
  summarize(ndsi_avg)
  
postsnow<-ndsi_avg %>%
  filter(year %in% postyears)%>%
  group_by(year) %>%
  summarize(ndsi_avg)

unburnedsnow<-ndsi_avg %>%
  filter(site %in% 'unburned')%>%
  group_by(site)%>%
  summarize(ndsi_avg)

burnedsnow<-ndsi_avg %>%
  filter(site %in% 'burned')%>%
  group_by(site)%>%
  summarize(ndsi_avg)

#Vegetation
preveg<-ndvi_avg %>%
  filter(year %in% preyears)%>%
  group_by(year) %>%
  summarize(ndvi_avg)

postveg<-ndvi_avg %>%
  filter(year %in% postyears)%>%
  group_by(year) %>%
  summarize(ndvi_avg)

unburnedveg<-ndvi_avg %>%
  filter(site %in% 'unburned')%>%
  group_by(site)%>%
  summarize(ndvi_avg)

burnedveg<-ndvi_avg %>%
  filter(site %in% 'burned')%>%
  group_by(site)%>%
  summarize(ndvi_avg)

#correlations
Pre<-cor(presnow$ndsi_avg,preveg$ndvi_avg)
Post<-cor(postsnow$ndsi_avg,postveg$ndvi_avg)
Unburned<-cor(unburnedsnow$ndsi_avg,unburnedveg$ndvi_avg)
Burned<-cor(burnedsnow$ndsi_avg,burnedveg$ndvi_avg)

#Answers in a table
answers<-data.frame(Pre,Post,Unburned,Burned)
answers

```
Based off our correlations, we see no relationship between ndsi vs ndvi when we consider pre and post fire, as well as unburned and burned. Other then Post, all other correlations are less than question 2. This alludes to other underlying variables impacting our values such as shifts in climate year to year, soil type, and runoff rates that could be influencing these relationships. 

## Question 4

What month is the greenest month on average? 

```{r, message=F}
green_avgmonth<-ndvi%>% 
  pivot_longer(
    cols= c("burned","unburned"),
    names_to = ("site"),
    values_to = ("value"),
    values_drop_na=TRUE) %>%
  mutate(month=month(DateTime))%>%
  group_by(month) %>%
  summarize(ndvi_avgmonth=mean(value)) %>%
  filter(ndvi_avgmonth == max(ndvi_avgmonth))

green_avgmonth


```

A:August is the greenest month on average.

## Question 5) 

What month is the snowiest on average?

```{r}
snow_avgmonth<-ndsi%>%
  pivot_longer(
    cols= c("burned","unburned"),
    names_to = ("site"),
    values_to = ("value"),
    values_drop_na=TRUE) %>%
  mutate(month=month(DateTime))%>%
  group_by(month) %>%
  summarize(ndsi_avgmonth=mean(value)) %>%
  filter(ndsi_avgmonth == max(ndsi_avgmonth))
```

A: January is the snowiest month on average. 

## Bonus Question: Redo all problems with `spread` and `gather` using modern tidyverse syntax.

## Bonus Question: Use Climage Engine to pull the same data for the assignment, but updated with 2020/2021 data.
