# LAGOS Spatial Analysis


```{r setup, include=FALSE}
library(tidyverse) # Tidy packages
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data
library(USAboundaries) #USA states and counties
library(ggthemes)
library(ggpubr)

```


First download and then specifically grab the locus (or site lat longs)

```{r data-read, warning=FALSE}
# #Lagos download script
LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path())

#Load in lagos
lagos <- lagosne_load()

#Grab the lake centroid info
lake_centers <- lagos$locus

```


Convert to spatial data
```{r 04 convert to spatial }
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

#st_as_sf, takes table and makes into spatial objects
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326) %>% #crs is projection system. diff codes
                          st_transform(2163) #transforming projection for map view

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
mapview(subset_spatial)

```

Subset to only Minnesota

```{r, 04 subsetting}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)
minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000) %>%
  mapview(.,zcol = 'lake_area_ha')
```

## Map

Show a map outline of Iowa and Illinois (similar to Minnesota map upstream).

```{r 04 mapping}
#filtering out the states of interest
Iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

Illinois <- states %>%
  filter(name == 'Illinois') %>%
  st_transform(2163)

#combining the two states and mapping them
maps <- rbind(Iowa,Illinois)
mapview(maps)
```


### Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

```{r 04 lake id}
iowa_lakes<-spatial_lakes[Iowa,]

illinois_lakes <- spatial_lakes[Illinois,]

```
A: There are 4,644 lakes in Iowa, and 11,822 lakes in Illinois for a combined total of 16,466 lakes. Minnesota has 29,038 lakes, almost double the amount in Iowa and Illinois. 

### What is the distribution of lake size in Iowa vs. Minnesota?

```{r}
#histogram plot with lake size on x-axis and frequency on y axis
iowa_size <- iowa_lakes %>%
  ggplot(aes(x=lake_area_ha))+
  geom_histogram(bins=20,
                 fill="cadetblue3",
                 color= "lightblue3",
                 alpha=0.5)+
  scale_x_log10()+
  theme_few()+
  labs(x= 'Area (ha)', y= 'Count',title="Iowa Lake Size Histogram")

minnesota_size <- minnesota_lakes %>%
  ggplot(aes(x=lake_area_ha))+
  geom_histogram(bins=20,
                 fill="cadetblue3",
                 color= "lightblue3",
                 alpha=0.5)+
  scale_x_log10()+
  theme_few()+
  labs(x= 'Area (ha)', y= 'Count',title="Minnesota Lake Size Histogram")

#presentig data
ggarrange(iowa_size,minnesota_size, ncol=1, nrow=2)
```

## Interactive Maps

Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r}
#combining datasets
Iowa_Illinois_lakes <- rbind(iowa_lakes,illinois_lakes) 

Iowa_Illinois_lakes %>%
  arrange(-lake_area_ha) %>%
  slice(1:1000) %>%
  mapview(.,zcol='lake_area_ha')

```


### What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

A: A dataset with lake depths would be beneficial when looking at lake size. This could potentially include the maximum, minimum, and average depth and would give us better information to the volume these lakes hold. 

